---
title: "SEMplR_Vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SEMplR_Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.height = 5,
  fig.width = 5
)
```

```{r}
suppressPackageStartupMessages({
  library(VariantAnnotation)
  library(SEMplR)
})
```

# Loading SEMs

First, we create a list of file paths to the .sem files we want to include. We will also load the meta data for each sem in a `data.table` object. If meta data is used, all SEMs must be represented in the meta data table.
```{r}
sem_files <- list.files("~/Downloads/SEMplR_2-14-25_freeze/SEMs/", 
                        pattern = ".sem", full.names = T)
metadata_file <- "~/Downloads/SEMplR_2-14-25_freeze/sempl_metadata.csv"
meta <- data.table::fread(file = metadata_file, sep = ",")

head(meta)
```

We will load the matrix and meta data for all SEMs in a single `SNPEffectMatrixCollection` object. This object has two slots, one containing a named list of the matrices and a second slot containing our meta data table with a key column connecting the meta data to the names of the matrices.
```{r}
sc <- loadSEMCollection(semFiles = sem_files, 
                        semMetaData = meta, 
                        semMetaKey = "SEM")
sc
```
Printing the `SNPEffectMatrixCollection`, we can see that we have 223 SEMs and 13 meta data features for each.

We can access all SEMs in collection the with the function `sems()` or some subset of matrices by specifying a vector of semIds in the `semId` parameter.
```{r}
sems(sc, semId = c("AP2B_HUMAN.SK-N-SH", "ZSCAN4_secondary"))
```

We can view the SEM meta data slot with the `semData` function.
```{r}
semData(sc)
```

# Build Variant Set

We will load the variants we want to score as a `VRanges` object.
```{r}
vr <- VRanges(seqnames = c("chr12", "chr19"),
              ranges = c(94136009, 10640062), 
              ref = c("G", "T"), alt = c("C", "A"))
vr
```

# Score Variants
```{r}
sempl_obj <- scoreVariants(vr = vr, semList = sc)
sempl_obj
```

```{r}
variants(sempl_obj)
```

```{r}
semData(sempl_obj)
```

```{r}
scores(sempl_obj)
```


# View Frame
```{r}
viewFrames(sempl_obj, vid = "chr12:94136009:G>C", sid = "AP2B_HUMAN.SK-N-SH")
```

```{r}
plotSEM(sc, motif = "AP2B_HUMAN.SK-N-SH", 
        motifSeq = "GCTTTGAGGC", highlight = 1)
plotSEM(sc, motif = "AP2B_HUMAN.SK-N-SH", 
        motifSeq = "GCTTTCAGGC", highlight = 1, hcol = "red")
```

```{r}
plotSemMotifs(sempl_obj, variant = "chr19:10640062:T>A")
```

```{r}
plotSemMotifs(sempl_obj, variant = "chr19:10640062:T>A",
              label = "transcription_factor")
```


```{r}
plotSemVariants(sempl_obj, semId = "TFAP4_HUMAN.HepG2")
```

